MLH1 Results Outline

### 2020-01-30 doc/Bivalent_Analysis.Rmd
updating the MLH1 Mixed model results

>>>>

#apply this before ploting, to display strains in desired order
#MLH1_data <- MLH1_data %>%
#  arrange(strain, sex, category, mouse) %>%
#  mutate(Original.Name = factor(Original.Name)) #another category that you want the order to match
#MLH1_data <- with(MLH1_data, MLH1_data[order(sex, strain),])

# sort your dataframe, by the focal categories
#MLH1_data$mouse <- factor(MLH1_data$mouse, levels=unique(MLH1_data$mouse))

### 2020-01-30 doc/Main_Report_Results.rmd
updating the MLH1 Mixed model results

@@

MLH1_data <- MLH1_data %>%
  arrange(strain, sex, category, mouse) %>%
  mutate(Original.Name = factor(Original.Name)) #another category that you want the order to match
MLH1_data <- with(MLH1_data, MLH1_data[order(sex, strain),])

@@

```{r rand, echo=FALSE, include=FALSE, eval=FALSE}
#str(AP_mouse_table_w.Ages)

@@

cor.test(AP_mouse_table.HQ.10plus$mean_co[AP_mouse_table.HQ.10plus$sex == "female"], 
    AP_mouse_table.HQ.10plus$var[AP_mouse_table.HQ.10plus$sex == "female"] )
```

# Main Figures

```{r table.for.plot, echo=FALSE}

@@

# A. Models for mean and variance of MLH1 foci per cell

@@

Initial results indicate that male MLH1 counts have less variance within subspecies but more between subspecies compared to female data.

```{r mixed.models.setup, echo=FALSE, warning=FALSE}
#https://rcompanion.org/handbook/G_03.html
#testing random effects within lme

#notes for testing random effects

DF.HetC.MixedModel <- subset(DF.HetC.MixedModel, select= -c(sex.x, sex.y,
                            subsp.y,subsp.x,
                    category.y, category.x,
                        strain.x,strain.y, mouse.1  )  )

#main df with variables coded as chr
DF.HetC.MixedModel <- DF.HetC.MixedModel
DF.HetC.MixedModel$subsp <- factor(DF.HetC.MixedModel$subsp, ordered = FALSE)
DF.HetC.MixedModel$sex <- factor(DF.HetC.MixedModel$sex, ordered = FALSE)
DF.HetC.MixedModel$strain <- factor(DF.HetC.MixedModel$strain, ordered = FALSE)
```

```{r HQ.dataset, echo=FALSE}

#also make a HQ dataset, tighter window for ages,

#mean_co.HQ  (this might remove some of the MOLF males)
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[DF.HetC.MixedModel$Ncells > 9,]
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[DF.HetC.MixedModel.HQ$age.weeks < 25,]
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[!(is.na(DF.HetC.MixedModel.HQ$mouse) | DF.HetC.MixedModel.HQ$mouse==""), ]
```

```{r Mol.dataset, echo=FALSE}
#mean_co.HQ.mol
DF.HetC.MixedModel.HQ.mol <- DF.HetC.MixedModel.HQ

DF.HetC.MixedModel.HQ.mol$subsp <- ifelse(grepl("MSM", DF.HetC.MixedModel.HQ.mol$strain), "Mol", 
        ifelse(grepl("MOLF", DF.HetC.MixedModel.HQ.mol$strain), "Mol", DF.HetC.MixedModel.HQ.mol$subsp))

DF.HetC.MixedModel.HQ.mol$subsp <- ifelse(grepl(1, DF.HetC.MixedModel.HQ.mol$subsp), "Dom", 
        ifelse(grepl(2, DF.HetC.MixedModel.HQ.mol$subsp), "Musc", DF.HetC.MixedModel.HQ.mol$subsp))

DF.HetC.MixedModel.HQ.mol <- DF.HetC.MixedModel.HQ.mol[!(is.na(DF.HetC.MixedModel.HQ.mol$mouse) | DF.HetC.MixedModel.HQ.mol$mouse==""), ]

@@

Currently I ran through 3 data sets: 1) base, 2) HQ and 3) Molf (from HQ)

`r kable(table(DF.HetC.MixedModel$sex, DF.HetC.MixedModel$strain))`

@@

subspecies levels: 2(3), Musc and Domestics (Mousiness)

strain levels: Don: 3, Musc: 4  (Mol: 2)

@@

#ranef(HetC.MixedModel.HQ)
#which is the test for random effects

@@

```{r MM.1.Mol, echo=FALSE}

HetC.MixedModel.HQ.mol <- lme(mean_co ~ subsp * sex, data=DF.HetC.MixedModel.HQ.mol, random=list(strain=pdDiag(~sex) ) )

as.data.frame( round(   summary(HetC.MixedModel.HQ.mol)$tTable, 6) )

@@

> MM.2.Mol

```{r MixedModel.2.mol, echo=FALSE}
#reduced models
Reduced.strain.HQ.mol <- lme(mean_co ~ subsp * sex, data=DF.HetC.MixedModel.HQ.mol, random = ~1|strain)
#summary(Reduced.strain.HQ)

as.data.frame(round(summary(Reduced.strain.HQ.mol)$tTable, 6))
#add a aption to this code chunk
#ranef(Reduced.strain.HQ.mol)

The model for MM.2.Mol has similar results as the above model, with just 2 subsps
This model has significant coefficients for sex and the 2 interaction effects (sex:musc, sex:mol)

> MM2.2

$$mouse .mean CO ~=~ subsp * sex  + \varepsilon$$

```{r MixedModel.2.2, echo=FALSE}
Reduced.all.HQ <- gls(mean_co ~ subsp * sex, data=DF.HetC.MixedModel.HQ)

as.data.frame( round(summary(Reduced.all.HQ)$tTable, 6))
#summary(Reduced.all.HQ)
#ranef(Reduced.all.HQ)
```

Above is the table of coefficients for MM.2.2, HQ data set, gls(), No random effects.  
I think this gls has sex and interaction term as significant.

@@

```{r MixedModel.3, echo=FALSE}
#strain = fixed
strain.fixed.full <- lm(mean_co ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)
#library(xtable)

#this table of results is broken -- might be too long -- lm
#lm has different names
as.data.frame(round( summary(strain.fixed.full)$coefficients, 6))

@@

full2noRandom <- anova(HetC.MixedModel.HQ, Reduced.all.HQ) # < .0001

strain2noRandom <- anova(Reduced.strain.HQ, Reduced.all.HQ) #NS
```

```{r MolossinusMixedModel, echo=FALSE, eval=FALSE, include=FALSE}
#base model with molossinus level ... why am I making molossinus model?
#mean.co
HetC.MixedModel.mol <- lme(mean_co ~ subsp * sex, data=DF.HetC.MixedModel.mol, random=list(strain=pdDiag(~sex) ) )

#mreducing the model (reducing the random effects)
Reduced.strain.mol <- lme(mean_co ~ subsp * sex, data=DF.HetC.MixedModel.mol, random = ~1|strain)

Reduced.all.mol <- gls(mean_co ~ subsp * sex, data=DF.HetC.MixedModel.mol)

@@

```{r Base.strain.fixed, echo=FALSE}

#make mixed model with strain as fixed, since there are interesting results

@@

@@

### Figures

@@

```{r variance.hist, echo=FALSE}

@@

```{r cv.plots, echo=FALSE}

@@

### modeling variance

@@

lme function, Cv modeled. Only sex is significant.

```{r cv.mod1.mol, echo=FALSE}

HetC.MixedModel.HQ_mol.cV_co <- lme(cV ~ subsp * sex, data=DF.HetC.MixedModel.HQ.mol, random=list(strain=pdDiag(~sex) ) )

as.data.frame( round(summary(HetC.MixedModel.HQ_mol.cV_co)$tTable, 6))

@@

```{r cV.models, echo=FALSE}
HQjust.strain.var_co <- lme(cV ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random = ~1|strain)

as.data.frame( round(summary(Reduced.all.HQ)$tTable, 6))

@@

Sex and sex*subsp interaction effect are significant.

```{r cV.mod.mol, echo=FALSE}

HQ.mol.cV_co <- lme(cV ~ subsp * sex, data=DF.HetC.MixedModel.HQ.mol, random = ~1|strain)

as.data.frame( round(summary(HQ.mol.cV_co)$tTable, 6))

Sex is significant, when mol is a seperate subspecies interaction is no longer significant.

$$mouse .cV CO ~=~ subsp * sex + \varepsilon$$
```{r more.cV.CO, echo=FALSE}
HQfull.Reduced.cv_co <- gls(cV ~ subsp * sex, data=DF.HetC.MixedModel.HQ)
as.data.frame( round(summary(HQfull.Reduced.cv_co)$tTable, 6))
#summary(HQfull.Reduced.cv_co)

#contrast models
full2strain.CV <- anova(HetC.MixedModel.HQ.var_co, HQjust.strain.var_co)#

full2noRandom.CV <- anova(HetC.MixedModel.HQ.var_co, HQfull.Reduced.cv_co)#

strain2noRandom.CV <- anova(HQjust.strain.var_co, HQfull.Reduced.cv_co)# 0.0469
```

$$mouse .var CO ~=~ subsp * sex + rand(strain*sex) + \varepsilon$$

```{r Base.Variance.MM, echo=FALSE, eval=FALSE}

#there's 1 mouse with just 1 cell remove
DF.HetC.MixedModel <- DF.HetC.MixedModel[!(is.na(DF.HetC.MixedModel$var) | DF.HetC.MixedModel$var==""), ]
DF.HetC.MixedModel$var <-  as.numeric(DF.HetC.MixedModel$var)

#main var.CO
HetC.MixedModel.var_co <- lme(var ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random=list(strain=pdDiag(~sex) ) )

as.data.frame( round(summary(HetC.MixedModel.var_co)$tTable, 6))

#summary(HetC.MixedModel.var_co)
#ranef(HetC.MixedModel.var_co)
```
Above error says missing values in object -- I don't understand. Mol version of data frame works...

Base DF, lme function, random sex specific strain effects

$$mouse .var CO ~=~ subsp * sex  + \varepsilon$$
```{r base.var.mol, echo=FALSE}
HetC.MixedModel.mol.var_co <- lme(var ~ subsp * sex, data=DF.HetC.MixedModel.HQ.mol, random=list(strain=pdDiag(~sex) ) )

@@

$$mouse .var CO ~=~ subsp * sex + rand(1|strain) + \varepsilon$$

```{r var.model.2, echo=FALSE, eval=FALSE}
just.strain.var_co <- lme(var ~ subsp * sex, data=DF.HetC.MixedModel, random = ~1|strain)

as.data.frame( round(summary(just.strain.var_co)$tTable, 6))

#summary(just.strain.cV_co)
```

lme model above

$$mouse .var CO ~=~ subsp * sex + \varepsilon$$

```{r var.mod.3, echo=FALSE, eval=FALSE}

full.Reduced.cv_co <- gls(cV ~ subsp * sex, data=DF.HetC.MixedModel)
#summary(full.Reduced.cv_co)

as.data.frame( (summary(full.Reduced.cv_co)$tTable))

gls model model above

```{r var.contrast, echo=FALSE, eval=FALSE}
#contrast models
full2strain.CV <- anova(HetC.MixedModel.varco, just.strain.var_co)#0.9997

@@

When Molossinus is coded as a separate subspecies, then interaction terms are no longer significant.

#### Pooled cells

For later analysis.

@@

Character assigned variables.  The values are now quite different. Male, sex coefficient is negative. The fixed sex coefficient is the sex effect independent of subspecies or strain.

## Proportion of chrm classes

@@

#this is a second mode of

@@

#write this table
write.table(OG.cURATE.table.mouse, "~./MLH1repo/data/CurrentCuratedDataset_13.1.20.csv", sep=",",
            row.names = TRUE)

@@

>>>>

load(file="C:/Users/alpeterson7/Documents/MLH1repo/data/CleanBivData_1.28.20.RData")

@@

#apply this before ploting
MLH1_data <- MLH1_data %>%
  arrange(strain, sex, category, mouse) %>%
  mutate(Original.Name = factor(Original.Name)) #another category that you want the order to match
MLH1_data <- with(MLH1_data, MLH1_data[order(sex, strain),])

@@

```{r rand.misc, echo=FALSE, include=FALSE, eval=FALSE}
#str(AP_mouse_table_w.Ages)

@@

cor.test(AP_mouse_table.HQ.10plus$mean_co[AP_mouse_table.HQ.10plus$sex == "female"], 
    AP_mouse_table.HQ.10plus$var[AP_mouse_table.HQ.10plus$sex == "female"] )

@@

# A. Models for av MLH1 counts foci per cell

ToDo: add the random effect tests!!

@@

@@

- Mol as a different subsp

- ( Nested strain in subsp?)

*The mouse average table used in this part excluded Q5 cells*

*the MLH1 mean is from mean - adj_MLH1 (males have +1 CO to count for PAR)*

```{r mixed.model.data.setup, echo=FALSE, warning=FALSE}
#https://rcompanion.org/handbook/G_03.html
#testing random effects within lme

#What are the main DFs for this code chunk
#AP_mouse_table_w.Ages   <- used to remove males which are too old (includes outgroups) 
#DF.HetC.MixedModel   <- only Mmusculus
#DF.HetC.MixedModel.HQ <- only mice of certain age, Q5 excluded

#remove the extra merged cols
DF.HetC.MixedModel <- subset(DF.HetC.MixedModel, select= -c(sex.y,sex.x,
                            subsp.y,subsp.x,
                    category.y,category.x, strain.x,
                    strain.y, mouse.1  )  )
#DONT NEED CATEGORY

#SWITCH TO HQ.DF
#mean_co.HQ  (this might remove some of the MOLF males)
#DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[DF.HetC.MixedModel$Ncells > 9,]  #this removes the MOLF female

table(DF.HetC.MixedModel$strain, DF.HetC.MixedModel$sex)

#afriad this is removing NA's
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[( (DF.HetC.MixedModel$age.weeks < 25) &  (DF.HetC.MixedModel$age.days != 13) ) | is.na(DF.HetC.MixedModel$age.weeks), ] #keeps mice without ages, and removes juvinilles

table(DF.HetC.MixedModel.HQ$strain, DF.HetC.MixedModel.HQ$sex)

#reformate Molosinus subsp
#mean_co.HQ.mol
DF.HetC.MixedModel.HQ$subsp <- ifelse(grepl("MSM", DF.HetC.MixedModel.HQ$strain), "Mol", 
        ifelse(grepl("MOLF", DF.HetC.MixedModel.HQ$strain), "Mol", DF.HetC.MixedModel.HQ$subsp))

# For some reason I changed subsp to nums (wtf)
DF.HetC.MixedModel.HQ$subsp <- ifelse(grepl(1, DF.HetC.MixedModel.HQ$subsp), "Dom", 
        ifelse(grepl(2, DF.HetC.MixedModel.HQ$subsp), "Musc", DF.HetC.MixedModel.HQ$subsp))

DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel.HQ[!(is.na(DF.HetC.MixedModel.HQ$mouse) | DF.HetC.MixedModel.HQ$mouse==""), ]

#main df with variables coded as chr
DF.HetC.MixedModel.HQ$subsp <- factor(DF.HetC.MixedModel.HQ$subsp, ordered = FALSE)
DF.HetC.MixedModel.HQ$sex <- factor(DF.HetC.MixedModel.HQ$sex, ordered = FALSE)
DF.HetC.MixedModel.HQ$strain <- factor(DF.HetC.MixedModel.HQ$strain, ordered = FALSE)
DF.HetC.MixedModel.HQ$species <- factor(DF.HetC.MixedModel.HQ$species, ordered = FALSE)

#kable(table(DF.HetC.MixedModel.HQ$sex, DF.HetC.MixedModel.HQ$strain))
```

This is the size of mice used within the Mixed model

`r kable(table(DF.HetC.MixedModel.HQ$sex, DF.HetC.MixedModel.HQ$strain))`

Can I indedate the average number of cells with go into each mouse.?

@@

subspecies levels: 2(3), Musc and Domestics and Molossinus

strain levels: Don: 3, Musc: 4  Mol: 2

@@

ranef(HetC.MixedModel.HQ)
#which is the test for random effects ?!

@@

@@

Can the above model be re-written with lmer so that random effect can be tested?

@@

```{r MixedModel.3, echo=FALSE}
#strain = fixed
lm.strain.fixed.full <- lm(mean_co ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)
#library(xtable)

#this table of results is broken -- might be too long -- lm
#lm has different names
as.data.frame(round( summary(lm.strain.fixed.full)$coefficients, 6))

@@

#full2noRandom <- anova(HetC.MixedModel.HQ, Reduced.all.HQ) # < .0001

#strain2noRandom <- anova(Reduced.strain.HQ, Reduced.all.HQ) #NS

@@

```{r Base.strain.fixed, echo=FALSE}
#make mixed model with strain as fixed, since there are interesting results
#Consider including these results from all FIXED MODEL

@@

DECIDE WHICH RESULTS TO DISPLAY.

@@

#### MLH1 variance Figures

Not sure why I have this section -- think this should be into a supplemental doc or something.

@@

```{r variance.hist, echo=FALSE, eval=FALSE}

@@

```{r cv.plots, echo=FALSE, eval=FALSE}

@@

### Modeling Variance

(use the same models are mean_co, )

1.	HetC.MixedModel.HQ <- lme(mean_co ~ subsp*sex, data=DF.HetC.MixedModel.HQ, random=list(strain=pdDiag(~sex) ) )

2.	Reduced.strain.HQ <- lme(mean_co ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random = ~1|strain)

3.	lm(mean_co ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)

4.	strain.fixed <- lm(mean_co ~ sex * strain, data=DF.HetC.MixedModel.HQ)

@@

@@

```{r cV.models, echo=FALSE}
HQjust.strain.var <- lme(var ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random = ~1|strain)

as.data.frame( round(summary(HQjust.strain.var)$tTable, 6))

@@

$$mouse .var CO ~=~ subsp * sex  + \varepsilon$$
```{r base.var.mol, echo=FALSE, eval=FALSE}
HetC.MixedModel.mol.var_co <- lme(var ~ subsp * sex, data=DF.HetC.MixedModel.HQ.mol, random=list(strain=pdDiag(~sex) ) )

@@

```{r var.contrast, echo=FALSE, eval=FALSE}
#contrast models
# review this logic again

full2strain.CV <- anova(HetC.MixedModel.varco, just.strain.var_co)#0.9997

@@

When Molossinus is coded as a separate subspecies, then interaction terms are no longer significant.  (subsp *sex )  interaction?

<!--
For later analysis.  Not sure what these are good for.

@@

Character assigned variables.  The values are now quite different. Male, sex coefficient is negative. The fixed sex coefficient is the sex effect independent of subspecies or strain.
-->

## Summary of Mixed Model Analyses

**mouse average MLH1 count**

Multiple models were tested to test for potential evo models -- for the mouse average CO counts.  (there were a mix of results -- and nuanced)

First full Mixed model -- no significant fixed factors --(the random effect might be significant).

The second model were strain was (nested and random), the fixed effects, sex and interaction of (subsp and sex), are much more significant. The coefficients indicate, males in general have 1 less in average, and the musc and molf subsp have ~3 more on average.

**Variance effect**

** Do these patterns hold across Quality bins? **

# Figure 1

Main figure, MLH1 strain avs on top, chrm proportion bar plot on bottom

```{r table.for.plot, echo=FALSE}
#why do I add in a different table?

#### Proportion of chrm classes

make a smaller propoertion chrm block --

@@

@@

#write this table
#write.table(OG.cURATE.table.mouse, "~./MLH1repo/data/CurrentCuratedDataset_13.1.20.csv", sep=",",
#            row.names = TRUE)

@@

# Trash / Extra

<!--

#this chunk migth need to be deleted // I think everything is already done within the BivData
#this is a second mode of

-->
